---
author: "Bethany Clark"
date: '2023-03-27'
output: html_document
---

<!--- This is an HTML comment in RMarkdown. You can use these comments to make notes that won't get read when running the code -->

<!--- {-} curly brackets hash means subsections won't get numbered -->

## Seaward Extensions 1: Create Background Raster

**What does this section cover:**

-   Creating a raster (grid) that can be used for delineation of a seaward extension boundary from a colony of breeding seabirds. This raster will:
  - Have a specified cell size (resolution)
  - Cover the location of the colony or colonies with a buffer greater than the maximum radius distance. 
  - Specify cells at either marine or terrestrial (ice can be classified as terrestrial.)

-   This is the required input for the tutorials on how to produce outputs for a single seabird colony (2) and multiple seabird colonies (3).

<!--- Add a space after the line to start the bulleted list -->

<!--- Placing **text** will make text bold -->

This analysis was performed in **`r sessionInfo()$R.version$version.string`**\

This document was last updated on **`r Sys.Date()`** <br>

<!--- In the code chunk below, we specify include = F, so that we will run the chunk but not include the chunk in the final document. We set a global argument in the code chunk of echo = T, so that in later code chunks, the code will be displayed in the RMarkdown document -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

**Load required R packages:**

If the packages fail to load, you will need to install the relevant packages.

```{r load-packages, include = TRUE}
## Load libraries ####
library(sf)
library(terra)
```

**Supply input data**

```{r}
## Colony data: location, abundance estimate
d <- read.csv("AdeliePenguin_example_dataset.csv")
head(d)

## Land polygon

## Maximum buffer distance (m)
buffer.dist <- 10000

## Specified resolution of grid cell (m)
spec_res <- 5000
  
```


