# Install R Packages

<!--- This is an HTML comment in RMarkdown. You can use these comments to make notes that won't get read when running the code -->

<!--- If you don't understand what a RMarkdown document is. Stop here. Go learn. -->

<!--- Equally. You must understand the difference between Markdown vs. RMarkdown -->

<!--- Remember, outside of the R code chunks we are now coding in HTML syntax, not R syntax -->

This chapter was built with **`r sessionInfo()$R.version$version.string`**\

The chapter was last updated on **`r Sys.Date()`** <br>

<!--- In the code chunk below, we specify include = F, so that we will run the chunk but not include the chunk in the final document. We set a global argument in the code chunk of echo = T, so that in later code chunks, the code will be displayed in the RMarkdown document -->

```{r track-intro-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
## we also specify in the options here to set the overall working directory
## back to the root directory of the R Project we are working in. We do this
## because by default , the working directory for R code chunks is the directory 
## that contains the Rmd document. We don't want this option given our file
## set up prefers the option of having the working directory specified as that
## where the R Project is.
knitr::opts_knit$set(root.dir = "..")
```

## Recommended R packages for use within the Marine Toolkit

To use many of the functions within the Marine Toolkit, you will need to install various R packages. This page outlines all the R packages and associated versions of the **key packages used throughout the toolkit**.

The page also provides support script to install all necessary R packages.

> If you do not succeed in installing a necessary R package using supporting code from this page, it may be because a specific R package requires some form of custom installation. Generally, you should be able to find a solution for installing the package on the internet. If you are still having problems installing a specific R package, then we recommend you contact the person who maintains the R package. You can typically find out who the package maintainer is by searching on the internet.

### Install the tidyverse: a primary R package for processing data

First, check if the `tidyverse` is installed or not. If it is not installed, then install it.

```{r}

# Define the package name
package_name <- "tidyverse"

# Check if the package namespace exists
if (!requireNamespace(package_name, quietly = TRUE)) {
  # Install the package
  install.packages(package_name)
  
  # Load the newly installed package
  library(package_name, character.only = TRUE)
} else {
  # Package is already installed, proceed with using it
  message(package_name, " is already installed.")
}


```

### Review & install packages needed in the toolkit

[After we build the entire toolkit, we must find a way of checking / listing all the names of the **KEY** packages we used and their associated versions. Below outlines a way of supporting people to test whether they have the right version of a package on their machine].

> [Will need to populate the dataframe packages.for.toolkit properly after we build the toolkit]

```{r}

packages.for.toolkit <- data.frame(package_names = c('dplyr',"track2KBA", "TESTpackage"),
                                   package_versions_toolkit = c("2.3.0", "1.0.5", "1.1.1"))

## Get information about all packages currently installed on your computer
installed_packages <- installed.packages()

## Create a new dataframe showing the names and versions of the packages installed
## on your machine
df.packages.machine <- data.frame(package_names = rownames(installed_packages),
                          package_versions_machine = installed_packages[, "Version"],
                          row.names = 1:nrow(installed_packages))

df.packages.machine

## create a dataframe comparing the KEY packages used for the toolkit compared to what
## you have or may need on your machine
df.package.compare <- left_join(packages.for.toolkit, df.packages.machine, by = "package_names") %>% 
  mutate(Packages_to_install = ifelse(is.na(package_versions_machine),
                                      "Package requires installation",
                                      "Package already installed"))

df.package.compare

## Review the packages that require installation
df.packages.to.install <- df.package.compare %>% dplyr::filter(Packages_to_install == "Package requires installation")

df.packages.to.install

## Install all necessary packages if these are not installed on your machine already
## NOTE: This option installs packages by name, not by name and version
install.packages(c(df.packages.to.install$package_names))

```

