---
author: "Bethany Clark"
date: '2023-03-27'
output: html_document
---

<!--- This is an HTML comment in RMarkdown. You can use these comments to make notes that won't get read when running the code -->

<!--- {-} curly brackets hash means subsections won't get numbered -->

## Seaward Extensions 1: Create Background Raster

This analysis was performed in **`r sessionInfo()$R.version$version.string`**\

This document was last updated on **`r Sys.Date()`** 
<br><br>

**What does this section cover:**

-   Creating a raster (grid) that can be used for delineation of a seaward extension boundary from a colony of breeding seabirds. This raster will:
    - Have a specified cell size (resolution).
    - Cover the location of the colony or colonies with a buffer greater than the maximum radius distance. 
    - Specify cells at either marine or terrestrial (ice can be classified as terrestrial).

-   This is the required input for the tutorials on how to produce outputs for a single seabird colony [link to 2] and multiple seabird colonies [link to 3].

<!--- Add a space after the line to start the bulleted list -->

<!--- Placing **text** will make text bold -->

<!--- In the code chunk below, we specify include = F, so that we will run the chunk but not include the chunk in the final document. We set a global argument in the code chunk of echo = T, so that in later code chunks, the code will be displayed in the RMarkdown document -->

```{r swe3-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
## we also specify in the options here to set the overall working directory
## back to the root directory of the R Project we are working in. We do this
## because by default , the working directory for R code chunks is the directory 
## that contains the Rmd document. We don't want this option given our file
## set up prefers the option of having the working directory specified as that
## where the R Project is.
knitr::opts_knit$set(root.dir = "..")

wd <- getwd()
```

**Load required R packages:**

If the packages fail to load, you will need to install them.

```{r swe3-load-packages, include = TRUE}
## Load libraries ####

#install.packages("sf")
#install.packages("terra")
library(sf)
library(terra)

```

**Supply input data**

[For Marine Toolkit, need to provide guidance on choice of resolution. Currently, 5km aligned with same scale as the Critchley et al. 2018 paper.]

The shapefile is not provide in the github repository as the file size was too large. The file can be downloaded here: https://data.bas.ac.uk/collections/e74543c0-4c4e-4b41-aa33-5bb2f67df389/

Alternatively, the output is provided, so this can be used directly in the next section.


```{r swe3-input}
## Colony data: location, abundance estimate
df <- read.csv(paste0(wd,"/data-input-files-bookdown/AdeliePenguin_example_dataset.csv"))
head(df)

## Land polygon
basemap <- read_sf(paste0(wd,"/data-input-files-bookdown/Coastline_high_res_polygon_v7.1"))

## Maximum colony radius distance (m)
max_colony_radius <- 149000

## Specified resolution of grid cell (m)
grid_res <- 5000
  
```

**Format colony locations as spatial objects**

```{r swe3-crs}
#Set coordinate reference system (CRS) for colony locations 
col_locs <- st_as_sf(df,coords = c("longitude","latitude"),
         crs = 4326) 

#If needed, convert colony data to basemap projection
if(st_crs(col_locs) != st_crs(basemap)){
  col_locs_proj <- st_transform(col_locs, crs = st_crs(basemap))
} else {
  col_locs_proj <- col_locs
}

#Plot to check
plot(st_geometry(col_locs_proj), pch = 19, col = "blue")

```

**Create raster**

```{r swe3-raster}
#First set a buffer (m) around the colony/colonies + the resolution of the grid to make sure the grid is large enough. 
raster_buffer <- max_colony_radius + grid_res

#Find the extent of the colony locations
bounds <- st_bbox(col_locs_proj)

#Create the blank raster grid with the crs of the land polygon
ras <- terra::rast(xmin = bounds[[1]] - raster_buffer, 
                   ymin = bounds[[2]] - raster_buffer, 
                   xmax = bounds[[3]] + raster_buffer, 
                   ymax = bounds[[4]] + raster_buffer, 
                   resolution = grid_res,
                   crs = st_crs(basemap)$wkt)


```

## Fill in raster values with land polygon basemap 

Convert the basemap, currently in polygon format, to a raster and then overlay with the blank raster

NOTE: Operation will take time depending on size of polygon data and detail of borders. i.e. Higher resolution maps take longer to process.

```{r swe3-land-mask}

basemap_vector <- vect(basemap)
mask <- terra::rasterize(basemap_vector, ras)
mask[is.na(mask)] <- 2
mask[mask == 1] <- NA
mask[mask == 2] <- 1
plot(mask)
#plot(st_geometry(col_locs_proj), pch = 19, col = "blue", add = T)

```

## Save the raster for use in the next section

```{r swe3-save}
terra::writeRaster(mask,paste0(wd,"/data-input-files-bookdown/seaward-ext-background-raster.tif"), overwrite = TRUE)
```

