# Tracking data: Load data into R and visualise

<!--- This is an HTML comment in RMarkdown. You can use these comments to make notes that won't get read when running the code -->

<!--- If you don't understand what a RMarkdown document is. Stop here. Go learn. -->

<!--- Equally. You must understand the difference between Markdown vs. RMarkdown -->

<!--- Remember, outside of the R code chunks we are now coding in HTML syntax, not R syntax -->

This tutorial uses example data from a project led by the BirdLife International partner in Croatia: BIOM

The citation for this data is: **TBC**

The example data can be downloaded from: **TBC - SBTD**

Analyses outlined in this chapter were performed in **`r sessionInfo()$R.version$version.string`**\

This chapter was last updated on **`r Sys.Date()`** <br>

<!--- In the code chunk below, we specify include = F, so that we will run the chunk but not include the chunk in the final document. We set a global argument in the code chunk of echo = T, so that in later code chunks, the code will be displayed in the RMarkdown document -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
## we also specify in the options here to set the overall working directory
## back to the root directory of the R Project we are working in. We do this
## because by default , the working directory for R code chunks is the directory 
## that contains the Rmd document. We don't want this option given our file
## set up prefers the option of having the working directory specified as that
## where the R Project is. By specifying double dots (or more), this is like saying
## go back one directory or more, as required.
knitr::opts_knit$set(root.dir = "..")
```

<br>

## Description of the example dataset

<!--- remember, single * is italics, ** is bold -->

Species tracked: Yelkouan Shearwater (*Puffinus yelkouan*)

Life-cycle stage when birds were tracked: chick-rearing

Site / source population birds tracked from: Lastovo SPA, Croatia

Years birds were tracked over: 2019, 2020

Devices birds were tracked with: GPS

Device model type: PathTrack nanoFix GPS/UHF transmitters (≤ 5.5 g)

[Figure showcasing Lastovo SPA and source populations]

<br>

## Load packages

**Load required R packages:**

If the package(s) fails to load, you will need to install the relevant package(s).

```{r load-packages, include = TRUE}

## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## Load libraries --------------------------------------------------------------
## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

## sf package for spatial data analyses (i.e. vector files such as points, lines, polygons)
library(sf)
## Tidyverse for data manipulation
library(tidyverse)
## ggplot2 for plotting opionts
library(ggplot2)
## rnaturalearth package for geographic basemaps in R
library(rnaturalearth)
## leaflet package for interactive maps in R
library(leaflet)
## lubridate for date time
library(lubridate)

```

<br>

## Storing, reading, and formatting raw tracking data

<br>

### Storing raw tracking data

The type of device you use will dictate what format your raw tracking data is stored in.

Typically, we will work with **.csv** files.

Good file management is critical when working with large tracking datasets.

[As a guide, the following file structure can support efficient data management]

<br>

### Reading raw tracking data into R / Rstudio

Depending on your file structure, type of raw data, and size of your overall data, we recommend reading data into R in a way that produces a single data frame (or tibble) for all your data required for a specific analysis.

[Example R code for reading in raw tracking data is provided in the Appendix]

<br>

### Format of data

Having data standardised into a common format greatly improves reproducible research, and also the ability for data to be used in other studies.

The primary format we recommend is that of BirdLife International's Seabird Tracking Database:
-   <https://www.seabirdtracking.org/>

We recognise, however, that this format may not be appropriate for all analyses. Nevertheless, we encourage users to standardise their data into a common format. This will facilitate the ease through which data can be reformatted when necessary for other analyses.

[Decide on best way to show example datasets - either as screen shot images? Or as example data files? Or perhaps as both. Maybe just taking subsets of the data as required.]

<br>

## Load raw tracking data

Below, we load the raw tracking data obtained for Yelkouan Shearwaters.

To see how this data was loaded into R originally, and merged to create a single data frame, see the example code in the Appendix.

The `load` function supports loading various R file formats. Here we are loading an `.Rdata` file. The file was previously saved with the name of `yelk`. So when we load the file, an object called `yelk` will be loaded into the working environment in R.

```{r load-data, include = TRUE}

## Load the example data for Yelkouan Shearwaters
load("data-testing/tracking-data/Tracking_YESH_raw.Rdata")

## view the first two rows of data
## First view the data in tibble format
head(yelk,2)
## Then view the data in data frame format
head(data.frame(yelk),2)

```

> **tibble vs data frame**: we don't go into the specifics of these different data formats. The key message is that each provides a different way of interacting with, or viewing, data. Both are essentially a mechanism through which to work with tabular data. (i.e. data in rows and columns)

<br>

## Format data to match that of the Seabird Tracking Database

In the example dataset, you will notice that the data is not in the format of that relating to the seabird tracking database.

We can reformat the data by extracting the relevant columns of information, and by adding in any information where it might be missing.


```{r sbtd-format, include=TRUE}

## First, add relevant columns of information to align with SBTD format
## the mutate functions allows you to add a new column of information.
## add the new columns and rename the object to a more standardised name.
df_sbtd <- yelk %>% dplyr::mutate(dataset_id = "tbc",
                                  scientific_name = "Puffinus yelkouan",
                                  common_name = "Yelkouan Shearwater",
                                  site_name = "Lastovo SPA",
                                  lat_colony = "tbc",
                                  lon_colony = "tbc",
                                  device = "GPS",
                                  age= "adult",
                                  sex= "unknown",
                                  breed_stage = "chick-rearing",
                                  breed_status = "breeding",
                                  argos_quality = NA,
                                  equinox = NA) 


## review the changes you have made (i.e. the new columns you have added)
head(data.frame(df_sbtd),2)

## Now select all the relevant columns to align data with the format of the 
## seabird tracking database.
## There are 21 columns of data in the format for the SBTD.
## Remember, when you use the select function, you can also rename columns simultaneously.
df_sbtd <- df_sbtd %>% dplyr::select(dataset_id,
                                     scientific_name,
                                     common_name,
                                     site_name,
                                     ## below for example, we select the column 
                                     ## called colony_code but rename it to colony_name
                                     colony_name = colony_code,
                                     lat_colony,
                                     lon_colony,
                                     device,
                                     bird_id = bird_id,
                                     track_id = bird_id,
                                     original_track_id = bird_id,
                                     age,
                                     sex,
                                     breed_stage,
                                     breed_status,
                                     date_gmt = dttm,
                                     time_gmt = dttm,
                                     latitude,
                                     longitude,
                                     argos_quality,
                                     equinox)

## review the changes again
head(data.frame(df_sbtd),2)
```

<br>

## Review of the example data so far

For the following columns, you may notice a few things: 
<br>

* dataset_id is specified as *tbc*. This is because until data has been loaded into the SBTD, it will not have a unique dataset identification code that would relate to the dataset stored in the SBTD.
* lat_colony, lon_colony are specified as *tbc*, because we still need to define what the colony coordinates would be for each of locations birds were tagged from.
* bird_id, track_id, original_track_id, are all specified with the same code. This is because when data is formatted to align with the format of the SBTD:
  * we have a code that relates to the bird that was tracked (bird_id)
  * we have a SBTD unique code that relates to each trip undertaken by the bird, when multiple trips are recorded (track_id). Note though, it is often the case that users do not provide data which has been pre-split into unique trips. Therefore, it is often the case that all entries relating to track_id match that of bird_id.
  * we have a user defined code that can relate to each trip undertaken by the bird. However, the same caveat in the case of track_id applies to this column of data too.
* argos_quality and equinox are both specified as NA. This is because our data relates to GPS data which does not have an argos_quality estimate (typical of PTT devices) or a measure relating to the equinox (typical of GLS devices).




```{r visualise-data, include=TRUE}
## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
## view all data ----
## ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

## convert data frame to sf spatial object
df_sbtd_plot <- st_as_sf(df_sbtd, coords = c("longitude", "latitude"), crs=4326) # 4326 = geographic WGS84

## number of datapoints
nrow(df_sbtd_plot)

## interactive plot
map.alldata <- leaflet() %>% ## start leaflet plot
  addProviderTiles(providers$Esri.WorldImagery, group = "World Imagery") %>% 
  ## plot the points. Note: leaflet automatically finds lon / lat colonies
  ## Colour accordingly.
  addCircleMarkers(data = df_sbtd_plot,
                   radius = 3,
                   fillColor = "cyan",
                   fillOpacity = 0.5, stroke = F) 

map.alldata



```
