---
author: "Bethany Clark"
date: '2023-03-27'
output: html_document
---

<!--- This is an HTML comment in RMarkdown. You can use these comments to make notes that won't get read when running the code -->

<!--- {-} curly brackets hash means subsections won't get numbered -->

## Seaward Extensions 1: Create Background Raster

**What does this section cover:**

-   Creating a raster (grid) that can be used for delineation of a seaward extension boundary from a colony of breeding seabirds. This raster will:
  - Have a specified cell size (resolution)
  - Cover the location of the colony or colonies with a buffer greater than the maximum radius distance. 
  - Specify cells at either marine or terrestrial (ice can be classified as terrestrial).

-   This is the required input for the tutorials on how to produce outputs for a single seabird colony [link to 2] and multiple seabird colonies [link to 3].

<!--- Add a space after the line to start the bulleted list -->

<!--- Placing **text** will make text bold -->

This analysis was performed in **`r sessionInfo()$R.version$version.string`**\

This document was last updated on **`r Sys.Date()`** <br>

<!--- In the code chunk below, we specify include = F, so that we will run the chunk but not include the chunk in the final document. We set a global argument in the code chunk of echo = T, so that in later code chunks, the code will be displayed in the RMarkdown document -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
## we also specify in the options here to set the overall working directory
## back to the root directory of the R Project we are working in. We do this
## because by default , the working directory for R code chunks is the directory 
## that contains the Rmd document. We don't want this option given our file
## set up prefers the option of having the working directory specified as that
## where the R Project is.
knitr::opts_knit$set(root.dir = "..")
```

**Load required R packages:**

If the packages fail to load, you will need to install the relevant packages.

```{r load-packages, include = TRUE}
## Load libraries ####
library(sf)
library(terra)
```

**Supply input data**

[For Marine Toolkit, need to provide guidance on choice of resolution. Currently, 5km aligned with same scale as the Critchley et al. 2018 paper.]


```{r}
## Colony data: location, abundance estimate
d <- read.csv("AdeliePenguin_example_dataset.csv")
head(d)

## Land polygon

## Maximum colony radius distance (m)
max_colony_radius <- 10000

## Specified resolution of grid cell (m)
grid_res <- 5000
  
```

**If needed, convert colony data to basemap projection**

[Can add an if statement]

```{r}

#need to convert to spatial
#col_locs_proj <- st_transform(d, crs = st_crs(basemap))

## Plot basemap and colonies to check
#plot(st_geometry(basemap)) 
#plot(st_geometry(col_locs_proj), add=T, pch = 19)

```

**Create raster**

```{r}
#First set a buffer (m) around the colony/colonies + the resolution of the grid to make sure the grid is large enough. 
raster_buffer <- max_colony_radius + grid_res

#Find the extent of the colony locations
bounds <- st_bbox(col_locs_proj)

#Create the blank raster grid with the crs of the land polygon
ras <- terra::rast(xmin = bounds[[1]] - raster_buffer, 
                   ymin = bounds[[2]] - raster_buffer, 
                   xmax = bounds[[3]] + raster_buffer, 
                   ymax = bounds[[4]] + raster_buffer, 
                   resolution = grid_res,
                   crs = st_crs(basemap)$wkt)

```

## Fill in raster values with land polygon basemap 

Convert the basemap, currently in polygon format, to a raster and then overlay with the blank raster

NOTE: Operation will take time depending on size of polygon data and detail of borders. i.e. Higher resolution maps take longer to process.

```{r}
basemap_vector <- vect(basemap)
mask <- terra::rasterize(basemap_vector, ras)
ras <- is.na(mask)
ras[ras==0] <- 2
plot(ras)

```

## Save the raster for use in the next session

```{r}
terra::writeRaster(ras, "./data-testing/Antarctica_5x5km/Antarctica_HighRes_5x5km_terra.tif",overwrite=T)
dir.create("./data-testing/Antarctica_1x1km/")
terra::writeRaster(ras,"./data-testing/Antarctica_1x1km/Antarctica_HighRes_1x1km_terra.tif",overwrite=T)
```

